---
title: "2016"
output: 
  html_document:
    theme: cerulean
---
<style type="text/css">
body { 
  font-size: 16px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, cache = F, fig.align = 'center')
```


```{r data}
library(tidyverse)

load("../results/state_results_16")
load("../results/p_clinton")
load("../results/ec_sims_16")
load("../results/ppc16")

actual_results <- read_csv("../data/state_results_16.csv")
```

# Model performance in 2016


```{r topline-ec}
clinton <- round(quantile(ec_sims_16[, 2], c(0.05, 0.95)))
trump <- round(quantile(ec_sims_16[, 1], c(0.05, 0.95)))
```

From the perspective of November 6th, 2016, Yapa gave Hillary Clinton a **`r paste0(round(mean(ec_sims_16[, 2] > ec_sims_16[, 1])*100), "%")`** chance of winning a majority of electoral college votes, and Donald Trump a **`r paste0(round(mean(ec_sims_16[, 1] > ec_sims_16[, 2])*100), "%")`** chance of winning a majority of electoral college votes.  Yapa aggregated 329 polls from 46 states in 2016 pulled from realclearpolitics.  Its predictions for 5 states (AL, DC, HI, ND, and WY) were pure prior.

```{r ecplot, results = T}
data_frame(
  electoral_votes = c(ec_sims_16[, 1], ec_sims_16[, 2]),
  candidate = c(rep("Trump", nrow(ec_sims_16)), rep("Clinton", nrow(ec_sims_16)))
) %>% filter(electoral_votes != 268) %>%
  ggplot() +
  aes(x = electoral_votes, fill = candidate, y = ..density..) +
  geom_histogram(alpha = 0.7, bins = 538,
                 position = "identity") +
  scale_x_continuous(limits = c(0, 538),
                     breaks = c(seq(0, 538, 100), 270)) +
  scale_fill_manual(values = c("blue", "red")) +
  geom_vline(xintercept = 270, lty = 2, col = "darkgrey") +
  theme_minimal() +
  labs(x = "electoral college votes", y = NULL, fill = NULL, 
       title = "distribution of potential electoral college outcomes") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 16, color = "#007ba7", hjust = 0.5),
        plot.subtitle = element_text(hjust = .5, color = "#007ba7", size = 10),
        strip.text = element_text(size = 15, hjust = .5,
                                  color = "#007ba7"),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(color = 'grey35'),
        axis.text.x = element_text(size = 11)) +
  facet_wrap(~candidate, nrow = 2) +
  guides(fill = F)

```


```{r table}
res_table <- state_results_16 %>% 
  mutate(pred_winner = ifelse(`Mean Clinton` > `Mean Trump`, 'Clinton', 'Trump')) %>%
  left_join(actual_results %>% rename(State = state)) %>%
  mutate(actual_winner = ifelse(dem > rep, 'Clinton', 'Trump')) %>%
  select(State, contains('Clinton'), `Result Clinton` = dem, 
         contains('Trump'), `Result Trump` = rep, `Predicted Winner` = pred_winner,
         `Actual Winner` = actual_winner) %>%
  mutate_if(is.numeric, function(x) paste0(round(x*100), "%")) %>%
  arrange(State)

calib <- mean(ppc16$dem > ppc16$lower & ppc16$dem < ppc16$upper)
```


Yapa 'called' the results in all but three states correctly.  It predicted that Hillary Clinton would win Michigan by 4% (she lost by 0.2%), Pennsylvania by 2% (she lost by 0.7%), and Wisconsin by 7% (she lost by 1.2%).  In all other states, the candidate who Yapa had winning _on average_, won.


```{r showtable}
DT::datatable(res_table, options = list(scrollX = TRUE, 
                                            scrollY = "200px",
                                            pageLength = 51))
```


```{r map, fig.height = 5, fig.width = 8, fig.align = 'center'}
library(maps)
library(plotly)

md <- map_data("state")

mapping_data <- md %>%
  left_join(p_clinton %>%
              mutate(region = tolower(state))) %>%
  left_join(actual_results %>%
              select(state, ev)) %>%
  left_join(data_frame(
              state = state.name, 
              abb = state.abb)) %>%
  mutate(state = paste0(state, "\nelectoral votes: ", ev)) %>%
  arrange(group, order) %>%
  select(state, lat, long, p_clinton, group, abb)


map_plot <- mapping_data %>%
  mutate(p_clinton = ifelse(p_clinton > 0.99, 0.99, p_clinton),
         p_clinton = ifelse(p_clinton < 0.01, 0.01, p_clinton),
         `p(clinton)` = round(p_clinton, 2)) %>%
  ggplot() +
  aes(x = long, y = lat, group = group, label = state, fill = `p(clinton)`) +
  geom_polygon(col = 'grey', lwd = 0.2, alpha = 0.75) +
  scale_fill_gradient2(low = 'red', midpoint = 0.5, mid = 'purple', high = 'blue') +
  guides(fill = F) +
  labs(title = NULL, x = NULL, y = NULL) +
  theme_void() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.border = element_blank()) 

ggplotly(map_plot) %>% 
  config(displayModeBar = FALSE) %>% 
  layout(xaxis = list(fixedrange = TRUE),
         yaxis = list(fixedrange = TRUE))

```

Yapa is well calibrated.  In `r paste0(round(calib*100), '%')` of states (`r paste0(round(calib*50))`/51), the actual result was within the 90% uncertainty interval for that state.

```{r ppc}
# Plot
ppc16 %>%
  ggplot() +
  aes(x = mean, y = dem, xmin = lower, xmax = upper, 
      label = abb, col = dem) + 
  geom_abline(intercept = 0, slope = 1, lty = 2, col = 'darkgrey') +
  geom_point(alpha = 0.8, cex = 0.7) +  
  geom_errorbarh(height = 0, alpha = 0.7, lwd = 0.4) +
  scale_x_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.25),
                     labels = paste0(seq(0, 1, 0.25)*100, "%")) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.25),
                     labels = paste0(seq(0, 1, 0.25)*100, "%")) +
  scale_color_gradient(low = "red", high = "blue") +
  guides(col = F) +
  theme_minimal() +
  labs(x = "prediction", y = "result",
       title = "Model predictions and uncertainty intervals vs election results: 2016") +
  theme(plot.title = element_text(size = 15, color = "#007ba7", hjust = 0.5),
        axis.title = element_text(color = "#007ba7"),
        axis.text = element_text(color = 'grey')) +
  annotate("text", x = 0.2, y = 0.8, label = "Clinton beat prediction",
           col = 'darkgrey') +
  annotate("text", x = 0.8, y = 0.2, label = "Trump beat prediction",
           col = 'darkgrey') 
```

