---
title: "2020 - House"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    toc_depth: 4
---
<style type="text/css">
body { 
  font-size: 16px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, cache = F, fig.align = 'center')
```


```{r data}
library(tidyverse)

load("../results/district_results")
load("../results/district_simulations")
load("../results/district_res_sims")
load("../results/district_results_dem")
load("../results/district_results_rep")
load("../results/district_p_dem")

district_ts <- read_csv("../results/district_ts.csv")
house_ts <- read_csv("../results/house_ts.csv")

```

_Updated `r gsub(" 0", " ", format(Sys.Date(), format="%B %d, %Y"))`, `r as.numeric(as.Date("2020-11-03") - Sys.Date())` days from election day 2020._

# Who will win control of the House?


```{r topline-house}
dem <- round(quantile(res_sims[, 2], c(0.05, 0.95)))
rep <- round(quantile(res_sims[, 1], c(0.05, 0.95)))
```

The Democrats are projected to win **between `r dem[1]` and `r dem[2]`** seats in the House.  The Republicans are projected to win **between `r rep[1]` and `r rep[2]`** seats in the House.  In `r dim(res_sims)[1]` simulations of the House elections from the perspective of `r gsub(" 0", " ", format(Sys.Date(), format="%B %d, %Y"))`, the Democrats win a majority in **`r paste0(round(mean(res_sims[, 2] > res_sims[, 1])*100), "%")`** and the Republicans win a majority in **`r paste0(round(mean(res_sims[, 1] > res_sims[, 2])*100), "%")`**.


```{r houseplot, results = T}
data_frame(
  seats = c(res_sims[, 1], res_sims[, 2]),
  candidate = c(rep("Republican", nrow(res_sims)), rep("Democrat", nrow(res_sims)))
) %>% 
  ggplot() +
  aes(x = seats, fill = candidate, y = ..density..) +
  geom_histogram(alpha = 0.7, bins = 180,
                 position = "identity") +
  scale_x_continuous(limits = c(120, 300),
                     breaks = c(seq(0, 435, 20))) +
  scale_fill_manual(values = c("blue", "red")) +
  geom_vline(xintercept = 218, lty = 2, col = "darkgrey") +
  theme_minimal() +
  labs(x = "seats in the House of Representatives", y = NULL, fill = NULL, 
       title = "distribution of potential House elections outcomes") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 16, color = "#007ba7", hjust = 0.5),
        plot.subtitle = element_text(hjust = .5, color = "#007ba7", size = 10),
        strip.text = element_text(size = 15, hjust = .5,
                                  color = "#007ba7"),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(color = 'grey35'),
        axis.text.x = element_text(size = 11)) +
  facet_wrap(~candidate, nrow = 2) +
  guides(fill = F)

```


```{r over time}
house_ts %>%
  gather(metric, value, -date) %>%
  mutate(party = sapply(strsplit(metric, "_"), tail, 1),
         party = ifelse(party == "dem", "Democrat", "Republican"),
         metric = sapply(strsplit(metric, "_"), head, 1)) %>%
  spread(metric, value) %>%
  ggplot() +
  aes(x = date, y = mean, ymin = lower, ymax = upper,
      col = party, fill = party) +
  geom_vline(xintercept = as.Date("2020-11-03"), lty = 2, col = "darkgrey") +
  geom_hline(yintercept = 218, lty = 2, col = "darkgrey") +
  geom_hline(yintercept = 199, lty = 9, col = "red", alpha = 0.7) +
  geom_hline(yintercept = 235, lty = 9, col = "blue", alpha = 0.7) +
  geom_line() +
  geom_ribbon(col = NA, alpha = 0.2) +
  scale_x_date(limits = c(as.Date("2020-07-01"), as.Date("2020-11-10")),
               date_breaks = "1 month",
               date_labels =  "%b, %Y") +
  scale_y_continuous(limits = c(0, 435),
                     breaks = c(0, 100, 200, 218, 400, 435)) + 
  theme_minimal() +
  scale_color_manual(values = c("blue", "red")) +
  scale_fill_manual(values = c("blue", "red")) +
  labs(col = NULL, fill = NULL, x = NULL, 
       title = 'how the House forecast has changed over time',
       y = "House seats") +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 9),
        plot.title = element_text(size = 16, color = "#007ba7", hjust = 0.5)) +
  guides(col = F, fill = F) 


```

```{r houseratings}
house_ratings <- p_dem %>%
  mutate(cat = case_when(p_dem < 0.25 ~ 'Safe Rep',
                         p_dem > 0.75 ~ 'Safe Dem',
                         p_dem >= 0.65 ~ "Likely Dem",
                         p_dem > 0.55 & p_dem < 0.65 ~ "Lean Dem",
                         p_dem > 0.35 & p_dem <= 0.45 ~ "Likely Rep", 
                         p_dem <= 0.35 ~ "Lean Rep", 
                         TRUE ~ "Uncertain")) 
```

# Who will win each district?

Among the 435 House districts, **`r length(house_ratings$district[!house_ratings$cat %in% c('Safe Dem', 'Safe Rep')])` (`r paste0(round(length(house_ratings$district[!house_ratings$cat %in% c('Safe Dem', 'Safe Rep')])/435*100), "%")`)** are competitive (meaning that the Democrat or Republican is not a shoe-in) as of `r gsub(" 0", " ", format(Sys.Date(), format="%B %d, %Y"))`.  The table below lists the competitive districts arranged by partisan lean.


```{r house_results, results = 'asis'}

house_ratings %>% 
  filter(p_dem > 0.25, p_dem < 0.75) %>% 
  group_by(cat) %>%
  arrange(-p_dem) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  mutate(rank = max(rank) - rank) %>%
  ggplot() +
  aes(x = reorder(cat, -p_dem), y = rank, label = district, col = p_dem) +
  geom_text(alpha = 0.8, size = 3, face = 'bold') +
  scale_x_discrete(position = "top") +
  theme_void() +
  theme(axis.text.x = element_text(size = 12, color = "#007ba7"),
        axis.text.y = element_blank(),
        plot.title = element_text(color = "#007ba7")) +
  labs(x = NULL, y = NULL) +
  guides(col = F) +
  scale_color_gradient(low = "red",  high = "blue")

```


```{r barplot, eval = F}
barplot_data <- house_ratings %>% 
  count(cat) %>% 
  mutate(cat = factor(cat, levels = c("Safe Rep", "Likely Rep", "Lean Rep", "Uncertain",  "Lean Dem", "Likely Dem", "Safe Dem" ))) %>%
  mutate(prop = n/sum(n),
         mid = 0.5*prop + cumsum(prop) - prop) %>%
  arrange(cat)
barplot_data %>%
  ggplot() + 
  aes(x = 1, y= prop, fill = cat) + 
  geom_bar(stat = 'identity', alpha = 0.8) + 
  coord_flip() + theme_void() + 
  scale_fill_manual(values = c("#cc0000", "#ff6666", "#ff9999", "grey", "#99ccff", "#6699ff", "#0000ff")) +
  guides(fill = F) 

```


```{r table}
DT::datatable(district_results, options = list(scrollX = TRUE, 
                                            scrollY = "200px",
                                            pageLength = 51))
```

# Competitive districts
```{r district-detail, results = TRUE}

plot_district_ts <- function(d) {
  plot_data <- district_ts %>% 
    filter(district == d) %>% select(-district) %>%
    gather(metric, value, -date) %>%
    mutate(party = sapply(strsplit(metric, "_"), tail, 1),
           metric = sapply(strsplit(metric, "_"), head, 1)) %>%
    spread(metric, value)
  current_rates <- district_ts %>% filter(district == d, date == max(date)) %>% select(mean_rep, mean_dem) %>% unlist()
  current_rates_pos <- district_ts %>% filter(district == d, date == max(date)) %>% select(mean_rep, mean_dem) %>% unlist()
  if(abs(current_rates_pos[1] - current_rates_pos[2]) < 0.03) {
    current_rates_pos[current_rates_pos == min(current_rates_pos)] <-  min(current_rates_pos) - 0.012
    current_rates_pos[current_rates_pos == max(current_rates_pos)] <- max(current_rates_pos) + 0.012
  }
  p <- plot_data %>%
    ggplot() +
    aes(x = date, y = mean, ymin = lower, ymax = upper,
        col = party, fill = party) +
  annotate("text", x = max(district_ts$date), y = current_rates_pos,
           label = paste0(round(current_rates*100), "%"),
           hjust = 0, col = c("red", "blue"), size = 3.5) +
    geom_line(aes(group = party)) +
    geom_ribbon(col = NA, alpha = 0.2) +
    scale_x_date(limits = c(as.Date("2020-07-01"), as.Date("2020-11-10")),
                 date_breaks = "1 month",
                 date_labels =  "%b, %Y") +
    scale_y_continuous(limits = c(0, 1),
                       breaks = seq(0, 1, 0.25),
                       labels = paste0(seq(0, 100, 25), "%")) +
    geom_vline(xintercept = as.Date("2020-11-03"), lty = 2, col = "darkgrey") +
    theme_minimal() +
    scale_color_manual(values = c("blue", "red")) +
    scale_fill_manual(values = c("blue", "red")) +
    labs(col = NULL, fill = NULL, y = NULL, x = NULL, title = NULL) +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 15, face = "bold",
                                    color = "#007ba7", hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.minor = element_blank()) +
    guides(col = F, fill = F)
  print(p)
}

p_win <- function(d, p = "dem") {
  oppo <- ifelse(p == "dem", "rep", "dem")
  res <- district_simulations %>%
    filter(district == d) %>%
    summarise(pwin = mean(value[party == p] > value[party == oppo])) %>%
    pull(pwin) %>%
    round(3)
  res
}

```


```{r districtts, results='asis'}
library(pander)
for (d in house_ratings$district[!house_ratings$cat %in% c('Safe Dem', 'Safe Rep')]) {
  pwin <- p_win(d, "dem")
  pwin <- case_when(pwin < 0.01 ~ "less than 1%",
                    pwin > 0.99 ~ "more than 99%", 
                    TRUE ~ paste0(round(pwin*100, 0), "%"))
  prep <- p_win(d, "rep")
  prep <- case_when(prep < 0.01 ~ "less than 1%",
                    prep > 0.99 ~ "more than 99%", 
                    TRUE ~ paste0(round(prep*100, 0), "%"))
   pander::pandoc.header(d, level = 2)
   pander::pandoc.p('\n')
   pander::pandoc.p(paste0('The Democrat is projected to receive **between ',
                           round(100*district_ts$lower_dem[district_ts$district == d 
                           & district_ts$date == max(district_ts$date)], 0), '% and ',
                           round(100*district_ts$upper_dem[district_ts$district == d 
                           & district_ts$date == max(district_ts$date)], 0), 
                           '%** of the vote in ', d, 
                           '. The Republican is projected to receive **between ',
                           round(100*district_ts$lower_rep[district_ts$district == d 
                           & district_ts$date == max(district_ts$date)], 0), '% and ',
                           round(100*district_ts$upper_rep[district_ts$district == d 
                           & district_ts$date == max(district_ts$date)], 0), 
                           '%**.  Out of ', dim(district_simulations)[1]/(435*3), 
                           ' simulations of the election in ', d, 
                           ', the Democrat wins **', pwin, '** and the Republican wins **', 
                           prep, '**.'))
   plot_district_ts(d)
   pander::pandoc.p('\n')
}
```

# Polls

```{r polls-table}
house_races <- read_csv('../data/house_races.csv')
house_races %>%
  arrange(state, district, days_out) %>%
  filter(Sample > 0) %>%
  mutate(Rep = paste0(round(rep/Sample, 2)*100, "%"),
         Dem = paste0(round(dem/Sample, 2)*100, "%"),
         Other = paste0(round(Other/Sample, 2)*100, "%")) %>%
  select(State = state, District = district, 
         Rep, Dem, Other, Sample, `Poll End Date` = end_date) %>%
  DT::datatable(options = list(scrollX = TRUE, scrollY = "500px", pageLength = nrow(house_races)))
```
